HiC Contact Map Comaprison Using Graphlet Approach

### 2018-04-30 papers/manuscripts/manuscript.tex
final draft of manuscript

Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge is that 
3D structure of a cell is not readily available. Based on
\cite{adhikari2016chromosome3d}, fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed 
between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures 
the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both 
inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can 
be found between normal cells and cancerous ones.

@@

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries we sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
Just as \cite{wang2013properties}, 
we used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.

@@

\section{results and discussions}
\subsection{Contact Map Orbit Vector Distance}
\begin{figure*}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_chr14.png}
    \end{subfigure}
    \caption{   
        \textbf{Pair-wise graphlet signature difference for
        all 276 contact maps:}
        Each point on a graph is the result of averaging all
        the distances across all loci of that contact map.
        ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
        \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).  
     }
    \label{fig:results_all}
\end{figure*}

Result of pair-wise contact map graphlet distances is
illustrated in supplementary materials. 
As an example, we plotted average
distances only for chromosome 
14 in Figure \ref{fig:orbit-distances_chr14}.
Each point on the graph is the
average of the graphlet distance vector 
of the two cell lines specified in the legend
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$).

@@

As is shown in
figure \ref{fig:orbit-distances_chr14} as well as t-test results
(to be discussed next in this paper),
we can compare pairs of contact maps in terms of their closeness
to each other. 
As an example, our results show that
for most inter-chromosomal contact maps, ALL
and RL cell lines are closer to each other than to other
cell lines. as  shown in 
\ref{fig:orbit_distances_chr14}, this observation is
true for the first 13 inter-chromosomal 
contact maps of chromosome 14 as well as it intra-chromosomal
contact map.
Although comparisons can be made for each contact map as to
whether which contact maps are more similar to each other,
no global pattern has been observed from graphlet distances.

It is worth mentioning that interchromosomal 
thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome 
on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_chr14} since they
would result in a bias towards zero in 
averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where 
average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes.

\subsection{MIC Comparison}
\begin{figure*}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_cancer_vs_normal.png}
    \caption{}
    \label{fig:contact_map_correlations_all}
    \end{subfigure}

@@

In addition to simple graphlet distance comparison,
we can compare graphlets by calculating
how often certain graphlets occur in them. By doing so
we measure the frequency distribution
of certain spatial structures in
each contact map. As a result of this approach
we can compare contact maps by calculating the 
correlation between their orbit distributions.
A higher correlation would mean higher similarity
in terms of spatial structure between the loci
involved.

This approach also allows for comparison of orbits
themsleves. By comparing average oribit correlations
across all contact maps of two cell lines, we can
have a measure of how similar certain orbits are
between cell lines. This can give us a 
\textit{global} measure
of how similar two cell lines are in terms of 
\textit{local} spatial positionings.

We caclulated pair-wise MIC values 
for each orbit in each of the 276
contact maps from MIT, ALL, RL, and CALL4 data separately. 
We found statistically significant difference between
normal vs cancer and cancer vs cancer correlations.
The results of staistical tests are shown in table
\ref{tab:statistical_results}. The difference
between the two cells are also illustrated in figure
\ref{fig:orbits_correlations}, which showes corrected
mean difference of MIC values. Figure 
\ref{fig:contact_map_correlations_all} shows the difference
between average pairwise
correlations. For most of the contact maps, correlation
between normal and cancer cells are smaller
than the correlation between cancer cells. This is corroborated
by the results of statical test which verify that such difference
is if fact signifcant.
Figures \ref{fig:orbits_correlations_normal_vs_cancer} and
\ref{fig:orbits_correlations_all} both show more details about
this difference in correlation. As can be observed, for obits
in $\mathbb{Q}$, normal correlations are smaller than cancer
correlations, while for the rest of the orbits, there is no
difference.

Figure \ref{fig:orbits_correlations} demonstrates that certain
orbits of Leukemic have higher correlation to each other 
than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.

\subsection{Statistical Analysis}
In order to qauntify our results, we have conducted one-sided
t-test in order to test whether the average pair-wise
MIC values across orbits are significantly different from 
each other and also from 1. 
However, pairwise correlation of cancer cells for
orbits in $\mathbb{Q}$ were significantly higher than
correlation between cancer cells and normal cells.
The results
for this test showed that all values are significantly less than 1;
A summary of can be found in
table \ref{tab:t-test}.
We have also conducted t-tests for 
pair-wise orbit signature vector distances across contact maps.
Please refer to supplementary material for result of the 
full list of t-test results.

>>>>

Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge is that 
3D structure of a cell is not readily available. 
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed 
between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures 
the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both 
inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can 
be found between normal cells and cancerous ones.

Recently, there has been many efforts at predicting the 3D conformation
of chromosomes using HiC data. 
\cite{noble2011three, rousseau2011three, hu2013bayesian, varoquaux2014statistical, trieu2014large, zhang20133d, lesne20143d, bau2011three, adhikari2016chromosome3d}.
These efforts, although different in
approach, usually translate 
interaction frequencies in contact maps as an inverse measure of 
distance. In particular \cite{adhikari2016chromosome3d} developed
a method, Chromosome3D which uses
distance geometry simulated annealing (DGSA) in order to
reconstruct genome structure. These reconstructed 3D 
structures could be used in order to compare normal and malignant
cell lines. However, the main obstacle to doing so is that the
process of reconstruction is computationally
expensive. Almost all reconstruction methods rely on either 
deterministic or heuristic optimization in order to find the best
confomation that matches contact maps. Also, there is no real 
way of verifying whether the reconstruction results are accurate.
Fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
\cite{gasser2002visualizing, lanctot2007dynamic, mateos2009spatially}
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.

@@

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries we sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
Just as \cite{wang2013properties}, 
we used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.
We also performed and extra normalization by
performing network deconvolution \cite{feizi2013network}
in order to remove
the effect of indirect interactions.

@@

\section{results and discussions}
\subsection{Contact Map Orbit Vector Distance}
\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit_distances_normal_vs_cancer.png}
        \caption{}
        \label{fig:orbit_distances_all_a}
    \end{subfigure}

    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit_distances_cells.png}
        \caption{}
        \label{fig:orbit_distances_all_b}
    \end{subfigure}

    \caption{   
        \textbf{Average Pair-wise graphlet signature difference 
        over all 276 contact maps:}
        Each point on a graph is the result of averaging all
        the distances across all loci of all contact maps for 
        each pair of cells. A 0.1 standard deviation error
        bar is also plotted for each point.
        ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}$).  
     }
    \label{fig:orbit_distances_all}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_chr14.png}
    \end{subfigure}
    \caption{   
     }
    \label{fig:orbit_distances_chr14}
\end{figure}

@@

We can compare pairs of contact maps in terms of their closeness
to each other. 
As an example, in figure
\ref{fig:orbit_distances_chr14}, all pairs of cells are
compared to each other in terms of their distance for contact
maps involving chromosome 14. We can se that 
for the first 13 inter-chromosomal 
contact maps, ALL and RL cell lines are closer to each other
than to other cell lines.

We performed one-way ANOVA statistical test to 
see if there are significant 
difference between cancer-normal and cancer-cancer pairs.
We found that the difference statistically significant difference.
($F(1, 1654) = 20.49, p < 0.0001$).
As illustrated in figure \ref{fig:orbit_distances_all_a},
we can see that normal-cancer cells have higher distance
from each other than cancer-cancer pairs.

We then continued to investigate each pair separately to
see if there is any significant difference between them.
Again, our statistical tests (ANOVA) showed significant difference
between pairs of cells.
The results are shown in figure \ref{fig:orbit_distances_all_b}.
We can see that ALL-RL pair are closest to each other while
ALL-MIT and MIT-RL are most distant. We found statistcally
significant difference for difference between individual
pairs except for ALL-CALL4 and CALL4-RL as well as
MIT-RL and ALL-MIT. The results of thse tests 
can be found in supplementary material.

The results in figure \ref{fig:orbit_distances_all_b} is also
in keeping with what we see in figure \ref{fig:orbit_distances_chr14}. For example, as mentioned earlier, our results show that
on average, ALL
and RL cell lines are closer to each other than to other
cell lines, which is also the case in figure
\ref{fig:orbit_distances_chr14} for majority of contact maps.

\subsection{MIC Comparison}

\begin{figure*}
    \centering
%    \begin{subfigure}[b]{\textwidth}
%        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_cancer_vs_normal.png}
%    \caption{}
%    \label{fig:contact_map_correlations_all}
%    \end{subfigure}

@@

In addition to comparing cells in terms of their
orbit distances, we can compare them by measuring how
often certain graphlets occur in their contact maps. By
doing so, we measure the frequency distribution of the spatial
structures represented by orbits in each contact map. In order
to see how closely such structures are distributed, we can
compare contact maps by calculating the correlation between
their orbit distributions. A higher correlation for certain orbits
would mean higher similarity in terms of that particular spatial
structure between the loci involved.

Before going on with the results,
it is worth mentioning that interchromosomal 
thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome 
on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
You can see the bias in 
figure \ref{fig:orbits_correlations} where 
average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between 
the same 
chromosomes.

We caclulated pair-wise MIC values 
for each orbit in each of the 276
contact maps from MIT, ALL, RL, and CALL4 data separately. 
We found statistically significant difference between
cancer-cancer and normal-cancer correlations.
($F(73, 1582)=6.29$, $p < 0.00001$, Wilk's $\Lambda = 0.775$)
 The difference  is also illustrated in figure
\ref{fig:orbits_correlations_normal_vs_cancer},
which plots corrected
mean difference of MIC values for cance-cancer and 
normal-cancer correlations.  
We performed statistical test to see if there is a 
significant difference correlation between individual pairs
of cells. Figure \ref{fig:orbits_correlations_all} demonstrate
such difference.
Average correlation over all contac maps
for normal-cancer pairs are smaller
than cancer-cancer pairs. This is corroborated
by the results of statical test which verify that such difference
is in fact signifcant.($F(365, 7882) = 3.91$, $p < 0.00001$, 
Wilk's $\Lambda = 0456$)
Figures \ref{fig:orbits_correlations_normal_vs_cancer} and
\ref{fig:orbits_correlations_all} both show more details about
this difference in correlation. As can be observed, for obits
in $\mathbb{Q}$, normal correlations are smaller than cancer
correlations, while for the rest of the orbits, there is no
difference.

Figure \ref{fig:orbits_correlations} demonstrates that certain
orbits of Leukemic have higher correlation to each other 
than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits NOT in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.
